---
title: "Ling 1269 Project - Checkpoint 5"
author: "Jarod Latta"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

toIPA = function(column) {
  col = gsub("dz", "dʒ", column)
  col = gsub("sh", "ʃ", col)
  col = gsub("ng", "ŋ", col)
  col = gsub("th", "θ", col)
  col = gsub("eu", "əʊ", col)
  col = gsub("uu", "ʌʊ", col)
  col = gsub("ou", "oʊ", col)
  col = gsub("au", "aʊ", col)
  col = gsub("/e", "ə", col)
  col = gsub("oi", "oɪ", col)
  col = gsub("æu", "æʊ", col)
  
  col
}

#getConstType = function() {}
```

# Checkpoint 5

## Loading & Aggregating Data

First we load in all of the data that collected from the podcast files.

```{r loaddata}
df2016 = as_tibble(read.csv("praat-coding/s1e01_29-09-2016.csv", stringsAsFactors = TRUE))
df2017 = as_tibble(read.csv("praat-coding/s3e01_22-08-2017.csv", stringsAsFactors = TRUE))
df2018 = as_tibble(read.csv("praat-coding/s5e01_03-07-2018.csv", stringsAsFactors = TRUE))
df2019 = as_tibble(read.csv("praat-coding/s3e01_22-08-2017.csv", stringsAsFactors = TRUE))
```

Now that with this data, we now need to stack these dataframes (without losing the date information).

```{r addDateConcat}
df2016 = df2016 %>% mutate(Year = "2016")
df2017 = df2017 %>% mutate(Year = "2017")
df2018 = df2018 %>% mutate(Year = "2018")
df2019 = df2019 %>% mutate(Year = "2019")
df = bind_rows(df2016, df2017) %>% bind_rows(df2018) %>% bind_rows(df2019)
```

```{r checkDf}
df
```

Now that we have all of the data read in, we can clean up the data. First thing's first, let's convert all of the pseudo-IPA coding that I did to make working in and exporting from Praat work into actual IPA.

```{r cleanIPA}
df = df %>% 
  mutate(PreConst = factor(toIPA(PreConst))) %>% 
  mutate(AftConst = factor(toIPA(AftConst))) %>% 
  mutate(StdVowel = factor(toIPA(StdVowel))) %>% 
  mutate(PresentVowel = factor(toIPA(PresentVowel)))
```

We will also want to fix any little weird typos I made while coding (such as capitalization differences for the same word).

```{r cleanCap}
df = df %>% 
  mutate(Word = factor(gsub("Episode", "episode", Word))) %>% 
  mutate(Word = factor(gsub("Hello", "hello", Word))) %>% 
  mutate(Word = factor(gsub("Open", "open", Word)))
```

## Plotting & Analyzing Data

### Understanding Canadian Raising

Before we look at speech production for Canadian Raising, we'll want to fully understand what that is. In the broadest sense, it is the stereotype of Canadian English Speakers saying "aboot" instead of "about". Now more scientifically, very few (if any body) actually says *"aboot"*, but the real change here is the vowel "æʊ" moving to be more like "ʌʊ" or even "əʊ". With this example, we can see the *raising* of the first part of the diphthong æ up to ʌ or ə. The other part of Canadian Raising involves the "oʊ" or "aʊ" vowel moving towards the same "ʌʊ" or "əʊ". This is less of an upward movement in the vowel space and more of a forward movement (even moving slightly down in the case of oʊ).

Let's look at the context where this phenomena occurs:

```{r}
levels(df$PreConst)
```

```{r raisingConstraints}
context = df %>% 
  select(PreConst, AftConst, ORaising, Stressed) %>% 
  mutate(PreType = factor(ifelse(PreType %in% c("ʃ", "s", "θ"), "voiceless fricative", ifelse(
    PreType %in% c("")
  ))))
```

### Exploring Raising

Now that our data is clean and we have our understanding of what Canadian Raising is, we'll want to analyze what we have starting with the question, does Barbara produce Canadian Raising more frequently than the other (non-Canadian English) speakers?

```{r rateOfRaising}
barbVsAll = df %>% 
  select(Speaker, ORaising, StdVowel, PresentVowel) %>% 
  mutate(Speaker = factor(ifelse(Speaker != "Barbara", "Other", "Barbara")))
#barbVsAll

barbVsAll %>%
  ggplot(aes(x=StdVowel, fill=ORaising)) +
  geom_bar(position="fill") +
  facet_wrap(vars(Speaker)) +
  coord_flip() +
  labs(title = "Rate of Canadian Raising by Standard Diphthong", subtitle = "Barbara vs All") +
  ylab(element_blank()) + xlab(element_blank()) + # remove axis labels bc unnecessary
  scale_fill_discrete(name="Raising") # change legend title
```

From this plot, we can see that Barbara exhibits raising significantly more than the other speakers. Note that this information does not take any year distinctions into account. Now we'll look at the difference of her Candadian Raising over the course of four years (2016-2019).

```{r plot}
barbVsTime = df %>% 
  select(Speaker, ORaising, StdVowel, PresentVowel, Year) %>% 
  filter(Speaker == "Barbara")
#barbVsTime

barbVsTime %>% 
  ggplot(aes(x=StdVowel, fill=ORaising)) +
  geom_bar(position="fill") +
  facet_wrap(vars(Year), nrow=1, ncol=4) +
  coord_flip() +
  labs(title = "Rate of Canadian Raising by Standard Diphthong", subtitle = "Barbara Over Time") +
  ylab(element_blank()) + xlab(element_blank()) + # remove redundant axes
  theme(axis.text.x = element_text(angle=45)) +
  scale_fill_discrete(name="Raising") # set legend title
```

Looking at this plot, we see some interesting things that I'm not fully able to explain. First, why do we have no tokens for "æʊ" in 2017 or 2019? This is simply a fluke. The sections of podcast that I coded for were lacking in this specific variable in those years. If this study were to be done on a larger scale, I would code for a greater amount of time and would (almost) certainly acquire tokens.

The next most interesting thing is the rate of raising for the "æʊ" value. Perhaps not ironically, this is also the ABOUT vowel that is used for the stereotype of Canadian Raising, so it is quite fascinating to see her rate of raising half. I have to concede that there might again be sample bias here. Barbara has lived in the United States for about 10 years, so for her to *still* be producing raising on the "æʊ" vowel in 2016 at a rate of 100% seems quite unlikely, especially seeing it drop to about 55% in 2018.